---
title: 'Playing with the Ant Picnic dataset'
author: Jose R. Martínez Batlle
date: "August, 2018"
header-includes:
  \usepackage{pdflscape}
  \newcommand{\blandscape}{\begin{landscape}}
  \newcommand{\elandscape}{\end{landscape}}
output:
  pdf_document
    # toc: true
    # toc_depth: 3
    # latex_engine: xelatex    
    # fig_caption: yes
  # lof: false
  # word_document: default
  # html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE)
```

## Ant Picnic
- Project leader: Dr. Magdalena Sorger, https://theantlife.com/ and Kristin Bedell
- Project URL: http://studentsdiscover.org/lesson/ant-picnic/
- Data analysis page: http://studentsdiscover.org/lesson/ant-picnic-data-analysis/
- Source data: https://codap.concord.org/releases/latest/static/dg/en/cert/index.html#shared=35710

## Packages and scripts

### Packages
```{r, message=FALSE, warning=FALSE, tidy=TRUE}
library(dplyr) #Organize data
library(ez) #ezCor function, compute an plot correlation matrix
library(devtools) #For R packages, R scripts
library(gclus) #For order.single function
library(ggplot2) #For plots
library(tidyr) #For data organizing
```

### Scripts
```{r, message=FALSE, warning=FALSE, tidy=TRUE}
scripts.others <- 'panelutils.R' #Correlation coefficients upper panel
ghsource.others <- 'https://raw.githubusercontent.com/JoeyBernhardt/NumericalEcology/master/'
invisible(
  source_url(
    paste0(
      ghsource.others,
      scripts.others)
    )
  )
```

## Reading the data
```{r readdata, tidy=FALSE}
#Reading source data. CSV file downloaded from
#https://codap.concord.org/releases/latest/static/dg/en/cert/index.html#shared=35710
tdir <- tempdir()
setwd(tdir)
ghsource <- 'https://raw.githubusercontent.com/geofis/antpicnic/master/'
fname <- 'antpicnic.csv'
download.file(paste0(ghsource, 'sampledata/', fname), fname)
d <- read.csv(fname)
str(d)
N <- nrow(d)
set.seed(131)
d[sample(1:N,10),1:4]
```
&nbsp;
## Selected variables
```{r selvars, tidy=FALSE}
d.sel <- d %>% select(
  Country:Bait_Type,
  Latitude:Temperature,
  -Habitat_type,
  -Date,
  Number_of_Ants)
set.seed(131)
d.sel[sample(1:N,10),1:4]
```
&nbsp;
## Environmental variables and number of ants, by country
```{r basics.country.bp, message=FALSE, warning=FALSE, tidy=FALSE, fig.width=8, fig.height=8}
basics.country.bp <- d.sel %>% gather(variable,value,-Country,-Biome,-Bait_Type) %>%
  mutate(Country=abbreviate(Country,8)) %>% 
  ggplot(aes(x=Country, y=value)) +
  geom_boxplot() +
  theme(
    text = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    ) +
  facet_wrap(variable~., scales = 'free_y', nrow=3)
```

\newpage
\newgeometry{margin=0.1in, top=0.1in, headheight=0.0in, footskip=0.2in, includehead, includefoot}
\blandscape
```{r basics.country.bp2, echo=FALSE, tidy=FALSE, out.width='0.95\\paperheight', fig.width=11, fig.height=8, fig.align='center'}
#Box-Plots of variables by country
basics.country.bp
```
\elandscape
\restoregeometry

&nbsp;
## Environmental variables and number of ants, by biome
```{r basics.biome.bp, message=FALSE, warning=FALSE, tidy=FALSE, fig.width=8, fig.height=8}
basics.biome.bp <- d.sel %>% gather(variable,value,-Country,-Biome,-Bait_Type) %>%
  ggplot(aes(x=Biome, y=value)) +
  geom_boxplot() +
  theme(
    text = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    ) +
  facet_wrap(variable~., scales = 'free_y', nrow=3)
```

\newpage
\newgeometry{margin=0.1in, top=0.1in, headheight=0.0in, footskip=0.2in, includehead, includefoot}
\blandscape
```{r basics.biome.bp2, echo=FALSE, tidy=FALSE, out.width='0.95\\paperheight', fig.width=11, fig.height=8, fig.align='center'}
#Box-Plots of variables by country
basics.biome.bp
```
\elandscape
\restoregeometry

## Number of ants, by bait type
```{r basics.bait.bp, message=FALSE, warning=FALSE, tidy=FALSE, fig.width=8}
basics.bait.bp <- d.sel %>%
  select(Bait_Type, Number_of_Ants) %>%
  mutate(Number_of_Ants_log1p = log1p(Number_of_Ants)) %>% 
  gather(variable,value,-Bait_Type) %>% 
  ggplot(aes(x=Bait_Type, y=value)) +
  geom_boxplot() +
  theme(
    text = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    ) +
  facet_wrap(variable~., scales = 'free_y', ncol = 2)
basics.bait.bp
```
&nbsp;
## Data frame for correlation analysis
```{r, tidy=FALSE}
d.fcor <- d.sel %>% select_if(is.numeric)
set.seed(131)
d.fcor[sample(1:N,10),1:5]
d.fcor.o <- order.single(cor(d.fcor)) #Ordered according to r value
```
&nbsp;
## Data frame for correlation analysis with transformed variables
```{r, tidy=FALSE}
d.fcor.l <- d.fcor %>%
  mutate(
    Latitude=abs(Latitude),
    Longitude=abs(Longitude)) %>%
  mutate_all(., funs(log1p))
set.seed(131)
d.fcor.l[sample(1:N,10),1:5]
d.fcor.l.o <- order.single(cor(d.fcor.l))
```
&nbsp;
## Correlation panel using ez package
```{r corpanez, message=FALSE, warning=FALSE, tidy=FALSE, fig.width=8, fig.height=6}
ezCor(
  d.fcor,
  r_size_lims = c(4,9),
  label_size = 3) +
  labs(title="Environmental variables correlation panel")
```
&nbsp;
## Correlation panel of raw variables using pairs function and panelutils.R script
```{r corpan, message=FALSE, warning=FALSE, tidy=FALSE, fig.width=8, fig.height=8}
#Pearson
pairs(d.fcor[,d.fcor.o],
      lower.panel = panel.smooth,
      upper.panel = panel.cor,
      diag.panel = panel.hist,
      main = "Pearson Correlation Panel")
#Spearman
pairs(d.fcor[,d.fcor.o],
      lower.panel = panel.smooth,
      upper.panel = panel.cor,
      diag.panel = panel.hist,
      method = 'spearman',
      main = "Spearman Correlation Panel")
```
&nbsp;
## Correlation panel of log-transformed variables using pairs function and panelutils.R script
```{r corpanl, message=FALSE, warning=FALSE, tidy=FALSE, fig.width=8, fig.height=8}
#Pearson
pairs(d.fcor.l[,d.fcor.l.o],
  lower.panel = panel.smooth,
  upper.panel = panel.cor,
  diag.panel = panel.hist,
  main = "Pearson Correlation Panel (log1p)")
#Spearman
pairs(d.fcor.l[,d.fcor.l.o],
      lower.panel = panel.smooth,
      upper.panel = panel.cor,
      diag.panel = panel.hist,
      method = 'spearman',
      main = "Spearman Correlation Panel (log1p)")
```
